---
title: "Managing mass spectrometry experiments"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Managing mass spectrometry experiments}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsExperiment}
    %\VignetteDepends{BiocStyle,rpx,Spectra}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsExperiment")`<br />
**Authors**: `r packageDescription("MsExperiment")[["Author"]] `<br />
**Last modified:** `r file.info("MsExperiment.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(MsExperiment)
library(Spectra)
library(BiocStyle)
```

# Introduction

The goal of the `MsExperiment` package is the store and handle all
data related to a mass spectrometry experiment. In this vignette, we
will describe how to create a `MsExperiment` object and populate it
with various types of data.

```{r}
library("MsExperiment")
```

# Getting data

We will use a small subset of the
[PXD022816](https://www.ebi.ac.uk/pride/archive/projects/PXD022816)
project [(Morgenstern et
al. (2020)](https://doi.org/10.1021/acs.jproteome.0c00956). The
acquisitions correspond to the Pierce Thermo HeLa digestion standard,
diluted to 50ng/uL with 97:3 + 0.1% formic acid, and acquired on a
QExactive instrument.

Below, we use the `r Biocpkg("rpx")` package to access the project
from the PRIDE repository, and download files of interest. Note that
these will automatically be cached in the `rpx` packages' cache
directory.

```{r}
library("rpx")
px <- PXDataset("PXD022816")
px
pxfiles(px)
```

The project provides the vendor raw files, the converted mzML files as
well as the identification mzid files. Let's download fractions 1 and
2 of the mzML files.

If you run these commands interactively and it's the first time you
use `pxget()`, you will be asked to create the `rpx` cache directory -
you can safelfy answer *yes*. The files will then be downloaded. Next
time you want to get the same files, they will be loaded automatically
from cache.

```{r}
(i <- grep(".+0[12].+mzML$", pxfiles(px), value = TRUE))
fls <- pxget(px, i)
fls
```

# MS Experiment object

Let's start by creating an empty `MsExperiment` object that we will
late populate with different pieces of data.

```{r}
msexp <- MsExperiment()
msexp
```

# Experiment files

Let's now start with our MS experiment management by saving the
relevant files in a dedicated `MsExperimentFiles` object. In addition
to the mzML files, let's also assume we have the human proteomics
fasta files ready. Later, when loading the data, we will refer the the
files in this `MsExperimentFiles` object.

```{r}
msfls <- MsExperimentFiles(mzmls = fls,
                           fasta = "homo_sapiens.fasta")
msfls
```

Let's add these files to the main experiment management object:

```{r}
experimentFiles(msexp) <- msfls
msexp
```

# Raw data

We can now create a `Spectra` object containing the raw data stored in
the mzML files. If you are not familiar with the `Spectra` object,
please refer to the [package
vignettes](https://rformassspectrometry.github.io/Spectra/articles/Spectra.html).

```{r}
sp <- Spectra(experimentFiles(msexp)[["mzmls"]])
sp
```

We can now add these the the main experiment management object:

```{r}
spectra(msexp) <- sp
msexp
```

# Third party applications

Let's now assume we want to search the spectra in our mzML files
against the `homo_sapiens.fasta` file. To do so, we would like to use
a search engine such as MSGF+, that is run using the command line and
generates mzid files.

The command to run MSGF+ would look like this (see the [manual
page](https://msgfplus.github.io/msgfplus/MSGFPlus.html) for details):

```
java -jar /path/to/MSGFPlus.jar \
     -s input.mzML \
     -o output.mzid
     -d proteins.fasta \
     -t 20ppm \ ## precursor mass tolerance
     -tda 1 \   ## search decoy database
     -m 0 \     ## fragmentation method as written in the spectrum or CID if no info
     -int 1     ## Orbitrap/FTICR/Lumos
```

We can easily build such a command for each of our input file:

```{r}
mzids <- sub("mzML", "mzid", basename(experimentFiles(msexp)[["mzmls"]]))
paste0("java -jar /path/to/MSGFPlus.jar",
       " -s ", experimentFiles(msexp)[["mzmls"]],
       " -o ", mzids,
       " -d ", experimentFiles(msexp)[["fasta"]],
       " -t 20ppm",
       " -m 0",
       " int 1")
```

Here, for the sake of time and portability, we will not actually run
MSGF+, but a simple shell script that will generate mzid files in a
temporary R directory.


```{r}
(output <- file.path(tempdir(), mzids))
cmd <- paste("touch", output)
cmd
```

The `cmd` variable holds the two commands to be run on the command
line that will generate the new files. We can run each of these
commands with the `system()` function.

```{r}
sapply(cmd, system)
```

Below, we recover the names of the created files (that are identical
to those in the `outputs` variable above) and add these files names to
our experiment.

```{r}
experimentFiles(msexp)[["mzids"]] <- mzids
experimentFiles(msexp)
msexp
```

We can also decide to store the commands that were used to generate
the mzid files in the experiment files' metadata slot. Here, we use
the convertion to name that metadata item `"mzmls_to_mzids"` to
document to input and output of these commands.

```{r}
metadata(experimentFiles(msexp))[["mzmls_to_mzids"]] <- cmd
metadata(experimentFiles(msexp))
```

The `existMsExperimentFiles()` can be used to check which files that
are associated with an experiment actually exist:

```{r}
existMsExperimentFiles(msexp)
```

# Conclusions

The `MsExperiment` object has been used to store files and data
pertaining to a mass spectrometry experiment. It is now possible to
save that object and reload it later to recover all data and metadata.


```{r}
saveRDS(msexp, "msexp.rda")
rm(list = ls())
```

```{r}
msexp <- readRDS("msexp.rda")
msexp
experimentFiles(msexp)
```

We can access the raw data as long as the mzML files that were used to
generate it still exist in their original location, which is the case
here as they were saved in the `rpx` cache directory.

```{r}
sp <- spectra(msexp)
sp
plotSpectra(sp[1000])
```


# Session information

```{r si}
sessionInfo()
```

# References
