---
title: "Managing mass spectrometry experiments"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Managing mass spectrometry experiments}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsExperiment}
    %\VignetteDepends{BiocStyle,rpx,Spectra}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsExperiment")`<br />
**Authors**: `r packageDescription("MsExperiment")[["Author"]] `<br />
**Last modified:** `r file.info("MsExperiment.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(MsExperiment)
library(Spectra)
library(BiocStyle)
```

# Introduction

The goal of the `MsExperiment` package is the store and handle all
data related to a mass spectrometry experiment. In this vignette, we
will describe how to create a `MsExperiment` and populate it with
various types of data.

```{r}
library("MsExperiment")
```

# Getting data

We will use a small subset of the
[PXD022816](https://www.ebi.ac.uk/pride/archive/projects/PXD022816)
project [@Morgenstern:2020]. The acquisitions correspond to the Pierce
Thermo HeLa digestion standard, diluted to 50ng/uL with 97:3 + 0.1%
formic acid, and acquired on a QExactive instrument.

Below, we use the `r Biocpkg("rpx")` package to access the project
from the PRIDE repository, and download files of interest. Note that
these will automatically be cached in the `rpx` packages' cache
directory.

```{r}
library("rpx")
px <- PXDataset("PXD022816")
px
pxfiles(px)
```

The project provides the vendor raw files, the converted mzML files as
well as the identification mzid files. Let's download fractions 1 and
2 of the mzML and mzid files.

If you run these commands interactively and it's the first time you
use `pxget()`, you will be asked to create the `rpx` cache directory -
you can safelfy answer *yes*. The files will then be downloaded. Next
time you want to get the same files, they will be loaded automatically
from cache.

```{r}
(i <- grep(".+0[12].+mzML$", pxfiles(px), value = TRUE))
fls <- pxget(px, i)
fls
```

# MS Experiment object

Let's start by creating an empty `MsExperiment` object that we will
late populate with different pieces of data.

```{r}
msexp <- MsExperiment()
msexp
```

# Experiment files

Let's now start with our MS experiment management by saving the
relevant files in a dedicated `MsExperimentFiles` object. In addition
to the mzML files, let's also assume we have the human proteomics
fasta files ready. Later, when loading the data, we will refer the the
files in this `MsExperimentFiles` object.

```{r}
msfls <- MsExperimentFiles(mzmls = fls,
                           fasta = "homo_sapiens.fasta")
msfls
```

Let's add these files to the main experiment management object:

```{r}
experimentFiles(msexp) <- msfls
msexp
```

# Raw data

We can now create a `Spectra` object containing the raw data stored in
the mzML files. If you are not familiar with the `Spectra` object,
please refer to the package vignettes.

```{r}
sp <- Spectra(experimentFiles(msexp)[["mzmls"]])
sp
```

We can now add these the the main experiment management object:

```{r}
spectra(msexp) <- sp
msexp
```


# Session information

```{r si}
sessionInfo()
```
